---
title: "Analisys of most harmful weather events in USA according to data of NOAA Storm Database"
author: "[Igor Goltsov](mailto:riversy@gmail.com)"
date: "June 20, 2015"
output: html_document
keep_md: true
---

## Synopsis

TODO: Prepare synopsis about completed research...

## Data Processing

The data for anaylis is provided by [NOAA][1]. The data about any storm events may be found there. But I will use the [package][2] that provided by Coursera [Reproducible Research][3] course. 

### Download package

I will check if *repdata_Fdata_FStormData.csv.bz2* file exists in my *data* folder and download the package from link shown below if it didn't.

```{r download_data_package,results='hide'}
package.name <- "data/repdata_Fdata_FStormData.csv.bz2"
package.url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

if (!file.exists(package.name)){
    download.file(package.url, package.name, method="curl")
}    
```

So now we can load it into the variable *storm.data* now and prepare it for further analisys. 

```{r load_data, cache=TRUE}
storm.data <- read.csv(bzfile(package.name))
dim(storm.data)
head(storm.data)
```

### Data Processing

Property and Crop Damage values should be arranged into one unit of measure

Function **arrange_unit_of_measure(value, exp)** will arrange value of damage for property and crop.

```{r, results='hide'}
arrange_unit_of_measure <- function(value, exp){
    value <- as.numeric(value)
    
    if (exp %in% c("k","K")){
        exp <- 3
    }
    
    if (exp %in% c("m","M")){
        exp <- 6
    }
    
    if (exp %in% c("b","B")){
        exp <- 9
    }
    
    if (is.numeric(value) && is.numeric(exp)){
        value*10^exp
    } else {
        0
    }
}
```

Use **apply()** to arrange units of measure for property damages.

```{r clarify_property_damage, cache=TRUE}
arranged.prop.damage <- as.numeric(
  apply(storm.data, 1, function(x) { 
    prop.dmg <- as.numeric(x[25])
    prop.dmg.exp <- x[26] 
    
    arrange_unit_of_measure(prop.dmg, prop.dmg.exp)
  })
)
summary(arranged.prop.damage)
```

Use **apply()** to arrange units of measure for property damages.

```{r clarify_crop_damage, cache=TRUE}
arranged.crop.damage <- as.numeric(
  apply(storm.data, 1, function(x) { 
    crop.dmg <- as.numeric(x[27])
    crop.dmg.exp <- x[28] 
    
    arrange_unit_of_measure(crop.dmg, crop.dmg.exp)
  })
)
summary(arranged.crop.damage)
```

Squeeze the data required for my analisys only.

```{r squeeze_necessary_data, cache=TRUE}
library(dplyr)
analisys.data <- select(
    storm.data,
    EVTYPE,
    FATALITIES,
    INJURIES
)

analisys.data$PROPERTY.DAMAGE <- arranged.prop.damage
analisys.data$CROP.DAMAGE <- arranged.crop.damage

names(analisys.data) <- c(
    "event_type",
    "fatalities",
    "injuries",
    "property_damage",
    "crop_damage"    
)

str(analisys.data)
```

Now the data is clean and ready for analisys. 

## Results

The most harmfull for population weather events could be identified by quantity of Injuries and Fatalities presented in the *analisys.data* variable. 

### Events most harmful with respect to population

The most harmful types of event are 20% of events with greatest population harm values.

```{r calculate_population_harm}
library(dplyr)
population_harm <- select(analisys.data, event_type, fatalities, injuries) %>% 
                    group_by(event_type) %>% 
                    summarise(harm = sum(fatalities + injuries))

quantiles <- quantile(
    population_harm[population_harm$harm > 0, ]$harm,
    probs = seq(0 , 1, 0.20)
)

print(quantiles)

limit_harm <- quantiles[[5]]

print(limit_harm)
```

It looks like all types of event with popiulation harm greatest then `r limit_harm` people are most harmful. Build the top of most harmful to population events. The events what cvalified as most harmful are called by name and all other events are called as "OTHER EVENTS". 

```{r fig.height=9, fig.width=8}
most_harmful_events <- select(population_harm, event_type, harm) %>%
                    filter(harm >= limit_harm) %>% 
                    arrange(desc(harm))

less_harmful_events <- select(population_harm, event_type, harm) %>%
                    filter(harm < limit_harm)

summary_of_harm <- rbind(
    most_harmful_events, 
    data.frame(
        event_type = "OTHER EVENTS",
        harm = sum(less_harmful_events$harm)
    )
)

# Print plot what illustrates most harmful events 

library(ggplot2)
ggplot(
        summary_of_harm[1:10, ], 
        aes(x="", fill=event_type, y=harm / 10^3)
    ) + 
    geom_bar(width=1, stat = "identity") +
    guides(fill=guide_legend(ncol=2)) +
    ggtitle("TOP 10 of most harmful events\nwith respect to population health") +
    xlab("") +
    ylab("Thousands of people affected") 
```

#### Complete list of weather events that most harmful to population 

```{r}
library(knitr)
names(summary_of_harm) <- c("Event Type", "People harmed")
kable(summary_of_harm, digits=2)
```

### Events that has the greatest economic consequences

Events that has the greatest economical consequences are 20% of events with greatest economical damage values.

```{r calculate_economical_damage}
library(dplyr)
economical_damage <- select(analisys.data, event_type, property_damage, crop_damage) %>% 
                    group_by(event_type) %>% 
                    summarise(damage = sum( (property_damage + crop_damage) / 10^6 ))

quantiles <- quantile(
    economical_damage[economical_damage$damage > 0, ]$damage,
    probs = seq(0 , 1, 0.20)
)

print(quantiles)

limit_damage <- quantiles[[5]]

print(limit_damage)
```

It looks like all types of event with damage greatest then `r limit_damage` millions of dollars are most harmful to economics. Build the top of events with greatest economic consequencies. The events what cvalified as most damagable are called by name and all other events are called as "OTHER EVENTS". 

```{r fig.height=9, fig.width=8}
most_damagable_events <- select(economical_damage, event_type, damage) %>%
                    filter(damage >= limit_damage) %>% 
                    mutate(damage = damage / 10^3) %>%
                    arrange(desc(damage))

less_damagable_events <- select(economical_damage, event_type, damage) %>%
                    filter(damage < limit_damage) %>%                
                    mutate(damage = damage / 10^3)
                    

summary_of_economical_damage <- rbind(
    most_damagable_events, 
    data.frame(
        event_type = "OTHER EVENTS",
        damage = sum(less_damagable_events$damage)
    )
)

library(ggplot2)
ggplot(
        summary_of_economical_damage[1:10, ], 
        aes(x="", fill=event_type, y=damage)
    ) + 
    geom_bar(width=1, stat = "identity") +
    guides(fill=guide_legend(ncol=2)) +
    ggtitle("TOP 10 of events that has\nthe greatest economical consequences") +
    xlab("") +
    ylab("Damage in billions of dollars") 
```


#### Complete list of weather events with greatest economical damage

```{r}
library(knitr)
names(summary_of_economical_damage) <- c("Event Type", "Damage in billions of dollars")
kable(summary_of_economical_damage, digits=2)
```


[1]:http://www.ncdc.noaa.gov/stormevents/
[2]:https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
[3]:https://class.coursera.org/repdata-015
