---
title: "Analisys of most harmful weather events in USA according to data of NOAA Storm Database"
author: "[Igor Goltsov](mailto:riversy@gmail.com)"
date: "June 20, 2015"
output: html_document
keep_md: true
---

## Synopsis

TODO: Prepare synopsis about completed research...

## Data Processing

The data for anaylis is provided by [NOAA][1]. The data about any storm events may be found there. But I will use the [package][2] that provided by Coursera [Reproducible Research][3] course. 

### Download package

I will check if *repdata_Fdata_FStormData.csv.bz2* file exists in my *data* folder and download the package from link shown below if it didn't.

```{r download_data_package,results='hide'}
package.name <- "data/repdata_Fdata_FStormData.csv.bz2"
package.url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

if (!file.exists(package.name)){
    download.file(package.url, package.name, method="curl")
}    
```

So now we can load it into the variable *storm.data* now and prepare it for further analisys. 

```{r load_data, cache=TRUE}
storm.data <- read.csv(bzfile(package.name))
dim(storm.data)
head(storm.data)
```

### Data Processing

Property and Crop Damage values should be arranged into one unit of measure

Function **arrange_unit_of_measure(value, exp)** will arrange value of damage for property and crop.

```{r, results='hide'}
arrange_unit_of_measure <- function(value, exp){
    value <- as.numeric(value)
    
    if (exp %in% c("k","K")){
        exp <- 3
    }
    
    if (exp %in% c("m","M")){
        exp <- 6
    }
    
    if (exp %in% c("b","B")){
        exp <- 9
    }
    
    if (is.numeric(value) && is.numeric(exp)){
        value*10^exp
    } else {
        NA
    }
}
```

Use **apply()** to arrange units of measure for property damages.

```{r clarify_property_damage, cache=TRUE}
arranged.prop.damage <- as.numeric(
  apply(storm.data, 1, function(x) { 
    prop.dmg <- as.numeric(x[25])
    prop.dmg.exp <- x[26] 
    
    arrange_unit_of_measure(prop.dmg, prop.dmg.exp)
  })
)
summary(arranged.prop.damage)
```

Use **apply()** to arrange units of measure for property damages.

```{r clarify_crop_damage, cache=TRUE}
arranged.crop.damage <- as.numeric(
  apply(storm.data, 1, function(x) { 
    crop.dmg <- as.numeric(x[27])
    crop.dmg.exp <- x[28] 
    
    arrange_unit_of_measure(crop.dmg, crop.dmg.exp)
  })
)
summary(arranged.crop.damage)
```

Squeeze the data required for my analisys only.

```{r squeeze_necessary_data}
library(dplyr)
analysis.data <- select(
    storm.data,
    EVTYPE,
    FATALITIES,
    INJURIES
)

analysis.data$PROPERTY.DAMAGE <- arranged.prop.damage
analysis.data$CROP.DAMAGE <- arranged.crop.damage

names(analysis.data) <- c(
    "EVENT.TYPE",
    "FATALITIES",
    "INJURIES",
    "PROPERTY.DAMAGE",
    "CROP.DAMAGE"    
)

str(analysis.data)
```

Now the data is clean and ready for analisys. 

## Results

The most harmfull for population weather events could be identified by quantity of Injuries and Fatalities presented in the *analysis.data* variable. 

### Most harmful for population









[1]:http://www.ncdc.noaa.gov/stormevents/
[2]:https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
[3]:https://class.coursera.org/repdata-015
